%%
% homework template
%


\documentclass{article}

%\usepackage{enumitem}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{epstopdf}
\usepackage{enumerate}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage{url}
\usepackage{mathtools}

\usepackage{dsfont}
\usepackage{tikz}
\usetikzlibrary{trees}
\usetikzlibrary{arrows,positioning,shapes.multipart}
\usepackage{makecell}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{listings}
\usepackage{bbm}
\usepackage{bm}
\usepackage{tikz}
\usetikzlibrary{trees}
\usepackage{makecell}
\usepackage{listings}
\usepackage{bbm}
\usepackage{bm}
\usepackage{xspace}

\usepackage{tcolorbox}
\tcbuselibrary{breakable} % to allow boxes to break across pages

% Define a new answer box environment with an optional height parameter.
\newtcolorbox{answerbox}[1][]{%
  breakable,
  colback=white,       % Background color
  colframe=black,      % Border color
  boxrule=0.5pt,       % Border thickness
  sharp corners,       % Box style
  before skip=10pt,    % Space before the box
  after skip=10pt,     % Space after the box
  % Allow adjustable height via an optional parameter (default: 3cm)
  height=\dimexpr\ifx#1\empty 3cm\else #1\fi\relax,
}

\usetikzlibrary{arrows,positioning,shapes.multipart}

\graphicspath{{./hw4/}}


\newcommand{\class}{ CS58700-DPL Spring 2025 }
\newcommand{\website}{{\tiny\url{https://www.cs.purdue.edu/homes/ribeirob/courses/Spring2025/index.html}}}
\newcommand{\homeworknumber}{4}
\newcommand{\duedate}{ {\bf 6:00pm}, April 20th (open until 6:00am April 21st)}
\newcommand{\code}{CODECODECODE}

%\pagestyle{headings}
\setlength{\parskip}{1pc}
\setlength{\parindent}{0pt}
\setlength{\topmargin}{-1pc}
\setlength{\textheight}{8.5in}
\setlength{\oddsidemargin}{0pc}
\setlength{\evensidemargin}{0pc}
\setlength{\textwidth}{6.5in}

\newcommand{\var}{\mbox{var}}
\newcommand{\cov}{\mbox{cov}}

\newcommand{\mA}{\bm{A}}
\newcommand{\mH}{\bm{H}}
\newcommand{\mW}{\bm{W}}
\newcommand{\mX}{\bm{X}}
\newcommand{\mY}{\bm{Y}}

\newcommand{\vb}{\bm{b}}
\newcommand{\vc}{\bm{c}}
\newcommand{\vh}{\bm{h}}
\newcommand{\vm}{\bm{m}}
\newcommand{\vx}{\bm{x}}
\newcommand{\vz}{\bm{z}}

\newcommand{\train}{\text{(tr)}}
\newcommand{\test}{\text{(te)}}
\newcommand{\source}{\text{(source)}}
\newcommand{\target}{\text{(target)}}


\newcommand*\dbar[1]{\overline{\overline{\lower0.2ex\hbox{$#1$}}}}
\newcommand{\harrow}[1]{\mathstrut\mkern2.5mu#1\mkern-11mu\raise1.6ex\hbox{$\scriptscriptstyle\rightharpoonup$}}

\newcommand{\newpart}{
\stepcounter{partno}
\noindent
{\bf (\alph{partno})}
}

\newcommand{\header}{
\newpage
\noindent
\framebox{ \vbox{\class Homework \hfill --- Homework \homeworknumber --- \hfill  Last update: \today
  \\ \website \hfill {\color{red} Due: \duedate} }}
\bigskip
\newline
{\bf Instructions and Policy:} Each student should write up their own solutions independently, no copying of any form is allowed.
You MUST indicate the names of the people you discussed a problem with; ideally you should
discuss with no more than two other people.\\
{\color{red} YOU MUST INCLUDE YOUR NAME IN THE HOMEWORK}\\
You need to submit your answer in PDF. {\LaTeX }  is typesetting is encouraged but not required.  Please write clearly and concisely - clarity and
brevity will be rewarded. Refer to known facts as necessary. 
\newline


}

\newcounter{questionno}
\setcounter{questionno}{-1}
\newcounter{partno}

\newcommand{\question}[1]{
\noindent
\newline
\stepcounter{questionno}
\setcounter{partno}{0}
{\bf Q\arabic{questionno} (#1 pts): }
}


%% -=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=- %%


\begin{document}

% Common homework head context.
%
\header
\question{%
    {\color{red}0pts correct answer, -1,000pts incorrect answer: (0,-1,000)}%
}
%
A correct answer to the following questions is worth 0pts.
%
An incorrect answer is worth -1,000pts, which carries over to other homeworks
and exams, and can result in an F grade in the course.
%
\begin{enumerate}[(1)]
\item Student interaction with other students / individuals:
\begin{enumerate}[(a)]
\item I have copied part of my homework from another student or another person (plagiarism).
\item \textbf{Yes, I discussed the homework with another person but came up with my own answers. Their name(s) is (are): Jason Fotso}
\item No, I did not discuss the homework with anyone
\end{enumerate}
\item On using online resources:
\begin{enumerate}[(a)]
\item I have copied one of my answers directly from a website (plagiarism).
\item \textbf{I have used online resources to help me answer this question, but I came up with my own answers (you are allowed to use online resources as long as the answer is your own). Here is a list of the websites I have used in this homework:\\ \url{https://pytorch.org/docs/main}}
\item I have not used any online resources except the ones provided in the course website.
\end{enumerate}

\end{enumerate}

\newpage

\section*{Homework \homeworknumber }

\input{concept}

\newpage

\input{program}



%############################################################
\newpage
\subsection*{Submission Instructions}
%
Please read the instructions carefully.
%
Failed to follow any part might incur some score deductions.

% Space.
%
\hfill

% PDF.
%
\subsection*{\bf PDF upload}
%
The report PDF must be uploaded on Gradescope.

% Space.
%
\hfill

% Code.
%
\subsection*{\bf Code upload}
%
\noindent \textbf{Naming convention}:
%
[firstname]\_[lastname]\_hw\homeworknumber
\\
%
All your submitting code files, a ReadMe, should be included in one folder. Create a folder [firstname]\_[lastname]\_hw\homeworknumber \ and put both the implemented skeleton folders inside it. Then tar it.\\
%
The folder should be named with the above naming convention.
%
For example, if my first name is ``Bruno'' and my last name is ``Ribeiro'',
then for Homework \homeworknumber, my file name should be
``bruno\_ribeiro\_hw\homeworknumber''.

% Space.
%
\hfill

% Pack your submission.
%
\noindent \textbf{Tar your folder:}
%
[firstname]\_[lastname]\_hw\homeworknumber.tar.gz
\\
%
Remove any unnecessary files in your folder, such as training datasets.
%
Make sure your folder structured as the tree shown in Overview section.
%
Compress your folder with the the command:
%
\textbf{
    tar -czvf bruno\_ribeiro\_hw\homeworknumber.tar.gz bruno\_ribeiro\_hw\homeworknumber
.}

% Space.
%
\hfill

% Submit.
%
\noindent \textbf{Submit:}
%
{\bf Brightspace}
\\
%
Please submit your compressed file on \textbf{Brightspace}.
%
Please make sure you didn't use any library/source explicitly forbidden to use.
%
If such library/source code is used, you will get 0 pt for the coding part of
the assignment.
%
If your code doesn't run on scholar.rcac.purdue.edu, then even if it compiles
in another computer, your code will still be considered not-running and the
respective part of the assignment will receive 0 pt.


\end{document}
